### 2018-06-21 paper/2-background.tex
Polishing from Simon
Esp early in section 5, and Excel in section 2

To support dynamic dependencies, \Excel's build
algorithm~\cite{excel_recalc} is significantly different from
\Make. \Excel{} arranges the cells into a linear sequence, called the
\emph{calc chain}.  During recalculation, \Excel{} processes
cells in the calc-chain sequence, but if computing a cell C requires the
value of a cell D that has not yet beeen computed, \Excel{}
\emph{aborts} computation of C, moves D before C in the calc chain,
and resumes recalculation starting with D.  When a calc cycle is
complete, the calc chain respects all the dynamic dependencies of the
spreadsheet.  When an input value (or formula) is changed, \Excel{}
uses the final calc chain from the \emph{previous} calc cycle as its
starting point so that, in the common case where changing an input
value does not change dependencies, there are no aborts.  Notice
that calc always succeeds regardless of the initial calc chain
(barring truly circular dependencies); the calc chain is just an optimisation.
We refer to this algorithm as \emph{restarting}, and will discuss it
in more detail in~\S\ref{sec-implementation-excel}.

>>>>

To support dynamic dependencies, \Excel's build
algorithm~\cite{excel_recalc} is significantly different from
\Make. \Excel{} arranges the cells into a linear sequence, called the
\emph{calculation chain}.  During recalculation, \Excel{} processes
cells the calc-chain sequence, but if computing a cell C requires the
value of a cell D that has not yet beeen computed, \Excel{}
\emph{aborts} computation of C, moves D before C in the calc chain,
and resumes recalculation starting with D.  When a calc cycle is
complete, the calc chain respects all the dynamic dependencies of the
spreadsheet.  When an input value (or formula) is changed, \Excel{}
uses the final calc chain from the \emph{previous} calc cycle as its
starting point so that, in the common case where chaning an input
value does not change dependencies, there are zero aborts.  Notice
that calc always succeeds regardless of the initial calc chain
(barring truly circular dependencies); the calc chain is ``just'' an optimisation.
We refer to this algorithm as \emph{restarting}, and will discuss it
in more detail in~\S\ref{sec-implementation-excel}.

### 2018-06-21 paper/3-abstractions.tex
Polishing from Simon
Esp early in section 5, and Excel in section 2

\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store
data Store i k v -- i = info, k = key, v = value
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
getHash    :: Hashable v => k -> Store i k v -> Hash v
initialise :: i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Hashing
hash :: Hashable a => a -> Hash a
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty
    Const x <*> Const y = Const (x <> y)
\end{minted}
\vspace{-3mm}
\caption{Signatures of main data types and library functions.}\label{fig-types}
\vspace{-4mm}
\end{figure}
We begin by establishing a common vocabulary for build systems:

@@

\noindent
Notice that these calls to \hs{dependencies} do no actual computation (in this
case, spreadsheet arithmetic). They cannot: we are not supplying a store or any
input numbers. So, through the wonders of polymorphism, we are able to extract
the dependencies of the spreadsheet formula, and to do so efficiently, simply by
running its code in a different \hs{Applicative}! This is not new, for example
see~\cite{free-applicatives}, but it is cool.

@@

\noindent
As expected, the dependencies of the cell \cmd{B1} from \hs{sprsh2} (see the
spreadsheet in~\S\ref{sec-task-monad}) are determined by the value of \cmd{C1},
which in this case is obtained by reading from the standard input.

>>>>

\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store
data Store i k v -- i = info, k = key, v = value
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
getHash    :: Hashable v => k -> Store i k v -> Hash v
initialise :: i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Hashing
hash :: Hashable a => a -> Hash a
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
modify    :: (s -> s) -> State s ()
runState  :: State s a -> s -> (a, s)
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty
    Const x <*> Const y = Const (x <> y)
\end{minted}
\vspace{-3mm}
\caption{Signatures of main data types and library functions.}\label{fig-types}
\vspace{-4mm}
\end{figure}
We begin by establishing a common vocabulary for build systems:

@@

\noindent
Notice that these calls to \hs{dependencies} do no actual computation (in this
case, spreadsheet arithmetic). They cannot: we are not supplying a store or any
input numbers. So, through the wonders of polymorphism, we are able to extract
the dependencies of the spreadsheet formula, and to do so efficiently, simply by
running its code in a different \hs{Applicative}! This is not new, for example
see~\cite{free-applicatives}, but it is etremely cool.  We will see a practical
use for \hs{dependencies} in \S\ref{sec-implementation-make}.

@@

\noindent
As expected, the dependencies of the cell \cmd{B1} from \hs{sprsh2} (see the
spreadsheet in~\S\ref{sec-task-monad}) are determined by the value of \cmd{C1},
which in this case is obtained by reading from the standard input.
We will see a use for \hs{track} in \S\ref{sec-implementation-shake}.

### 2018-06-21 paper/5-implementations.tex
Polishing from Simon
Esp early in section 5, and Excel in section 2

These two abstractions are the key to modularity: \emph{we can combine any scheduler
  with any rebuilder, and obtain a correct build system}.
In this section we will write a scheduler for each column of Table~\ref{tab-build-systems},
and a rebuilder for each row; then combine them to obtain the build systems in the
body of the Table.

@@

An implementation of \Make using our framework is shown in
Fig.~\ref{fig-make-implementation}. As promised, its definition
is just the application of a \hs{Scheduler}, \hs{topological},
to a \hs{Rebuilder}, \hs{modTimeRebuilder}.
We discuss each component in turn, starting with the rebuilder.

The \hs{modTimeRebuilder} uses the pair
\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
build information, carried by a \hs{MonadState} monad. This
\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
modification times}. The rebuilder receives three arguments: a
\hs{key}, its current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
the \hs{key} if necessary.  The rebuilder first decides if the
\hs{key} is \hs{dirty} by consulting \hs{modTime}: if the key is not
found, that must mean it has never been built before; otherwise
\hs{modTimeRebuilder} can see if any of the task's dependencies
(computed by \hs{dependencies}) are out of date.  If the key is
\hs{dirty}, we use `run task` to rebuild it, and update the state with
the new modification time of the key\footnote{The real \hs{make}
relies on the file system to track modification times, but we prefer
to make it explicit here.}; othewise we can just return the old vale.

>>>>

We write a scheduler for each column of Table~\ref{tab-build-systems}, and a
rebuilder for each row; then applying the former to the latter, in any combination,
gives a correct build system.

@@

An implementation of \Make using our framework is shown in
Fig.~\ref{fig-make-implementation}. As promised, its definition
is just the application of a \hs{Scheduler}, \hs{topological},
to a \hs{Rebuilder}, \hs{modTimeRebuilder}.
of the scheduler-rebuilder decomposition in Table~\ref{tab-build-systems}. We discuss
each component in turn, starting with the rebuilder.

The rebuilder \hs{modTimeRebuilder} uses the pair
\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
build information, carried by a \hs{MonadState} monad. This
\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
modification times}. The rebuilder receives three arguments: a
\hs{key}, its current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
the \hs{key} if necessary.  The rebuilder first decides if the
\hs{key} is \hs{dirty} by consulting \hs{modTime}: if the key is not
found, that must mean it has never been built before; otherwise
\hs{modTimeRebuilder} can see if any of the task's dependencies
(computed by \hs{dependencies}) are out of date.  If the key is
\hs{dirty}, we use `run task` to rebuild it, and update the state with
the new modification time of the key\footnote{The real \hs{make}
relies on the file system to track modification times, but we prefer
to make it explicit here.}; othewise we can just return the old vale.
